Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-04-17 closed_loop_planning.tex
[MGa] textual description of closed-loop planning

Building upon our previous paper that presented a 3D multibody model suitable for embedding within an MLTP problem~\cite{Domenighini:Designs:2023} - treated here as the \emph{baseline} model - two models with increasing complexity are presented. The first represents a vehicle capable of independently steering all four wheels (AS) and applying four independent torques to each wheel (AIWD, aka, torque vectoring), whereas the second model additionally incorporates an active camber variation system for each wheel (DCC).
%The kinematic tree of the two models is slightly different, but the procedure used to build the dynamic model is similar apart from a non-trivial aspect that will be discussed later on. In the remainder of this study, unless otherwise specified, the descriptions provided apply to both models. The figures related to Sections~\ref{sec:suspanal}-\ref{sec:tyremodel} are courtesy of the authors.

%% CHANGE THESE SECTIONS TO MERGE BOTH MODELS DESCRIPTIONS
\subsection{Kinematic Tree Structure of the Vehicle}
\begin{figure}[b]\centering
	\begin{tikzpicture}[scale=1.2]\small
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw,gray](g)at(-0.1,-0.1){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](b)at(0,2){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](h11)at(-0.9,2.9){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](h12)at(0.9,2.95){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](h21)at(-1,1){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](h22)at(0.95,1){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](k11)at(-2.0,3){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](k12)at(2.0,3){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](k21)at(-2.1,1.1){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](k22)at(2.0,1.05){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](w11)at(-3.0,2.9){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](w12)at(3.0,2.9){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](w21)at(-3.1,1.){};
		\node[fill,circle,minimum size=3pt,inner sep=0pt,draw](w22)at(3.0,.95){};
		\draw[gray](g)to node(2){}(b);
		\draw(b)to node(4){}(h11);
		\draw(b)to(h12);
		\draw(b)to(h21);
		\draw(b)to(h22);
		\draw(h11)to node(8){}(k11);
		\draw(h12)to(k12);
		\draw(h21)to(k21);
		\draw(h22)to(k22);
		\draw(k11)to node(6){}(w11);
		\draw(k12)to(w12);
		\draw(k21)to(w21);
		\draw(k22)to(w22);
		\draw(g)node[below,gray](1){$\{G\}$}(b)+(0,0.1)node[above](3){$\{B\}$}(h11)node[above](5){$\{H_1\}$}(h12)node[above]{$\{H_2\}$}(h21)node[below]{$\{H_3\}$}(h22)node[below]{$\{H_4\}$}
		(k11)node[above](9){$\{K_1\}$}(k12)node[above]{$\{K_2\}$}(k21)node[below]{$\{K_3\}$}(k22)node[below]{$\{K_4\}$}
		(w11)node[above](7){$\{W_1\}$}(w12)node[above]{$\{W_2\}$}(w21)node[below]{$\{W_3\}$}(w22)node[below]{$\{W_4\}$};
		\node[gray,inner sep=1pt](world)at(1,-0.3){world};
		\node[gray,inner sep=0pt,align=center](6DoF)at(-1.25,0.15){floating base\\(virtual 6-DoF)};
		\node[inner sep=1pt](chassis)at(0.05,3){chassis};
		\node[inner sep=0pt,align=center](susp)at(-2.2,1.6){suspension linkages\\(generalized 1-DoF)};
		\node[inner sep=1pt](knuckle1)at(-.4,3.7){knuckles 1};
		\node[inner sep=1pt](knuckle2)at(-2.05,3.8){knuckles 2};
		\node[inner sep=0pt,align=center](hub)at(-4.5,2.25){hub bearings\\(revolute 1-DoF)};
		\node[inner sep=0pt,align=center](camber)at(-2.,2.4){camber joint\\(revolute 1-DoF)};
		\node[inner sep=1pt](rim)at(-3.4,3.65){rims};
		\draw[gray](world.west)to(1.east);
		\draw[gray](6DoF.north east)to(2);
		\draw(chassis.south)to(3);
		\draw(susp.north east)to(4);
		\draw(knuckle1.south)to(5);
		\draw(knuckle2.south)to(9);
		\draw(hub.east)to(6);
		\draw(camber.north)to(8);
		\draw(rim.south east)to(7);
	\end{tikzpicture}
	\caption{Kinematic tree representing the interconnections between bodies. Bodies are referred to by their attached frame.}
	\label{fig:tree}
\end{figure}
The vehicle is modeled as an articulated rigid-body system with chassis, knuckles, and wheels as rigid bodies. The chassis represents all sprung masses rigidly connected, while the knuckles (zero-mass bodies) decouple steering and vertical suspension motion from the wheel spindle. In the second model, knuckles are split into two zero-mass bodies, introducing an additional degree of freedom per wheel to simulate dynamic camber control (DCC). Wheels include all rotating unsprung masses, with non-rotating parts (e.g., brake calipers) incorporated into the wheel-tire mass.

Each body has a right-handed barycentric frame: $\sref{B}$ (chassis), $\sref{H}$ and $\sref{K}$ (knuckles), and $\sref{W}$ (wheels). Additional frames $\sref{G}$ (ground) and $\sref{S}$ (centerline) are defined, with \(x\)-axis of $\sref{S}$ tangent to the centerline, pointing forward. Frame relationships are described by homogeneous transformation matrices $ g(d, \vPhi) \in SE(3) $, with $ d $ as position and \(\vPhi\) (Euler angles) parameterizing the rotation.

As shown in Fig.~\ref{fig:tree}, bodies are connected by joints in a kinematic tree. The second model generalizes the first, where $\sref{H}$ and $\sref{K}$ coincide, so only the second model's kinematic tree is shown for brevity.
%% START ORIGINAL
%The vehicle is modeled as an articulated rigid-body system, with chassis, knuckles, and wheels treated as rigid bodies. The chassis includes all sprung masses, rigidly connected, while the knuckles (modeled as zero-mass bodies) decouple steering and vertical suspension motion from the wheel spindle. In the second model, the knuckles are split into two zero-mass bodies, adding one degree of freedom per wheel to simulate dynamic camber control (DCC). The wheels encompass all rotating unsprung masses, with non-rotating parts (e.g., brake calipers) included in the wheel-tire mass.
%
%Each body has a right-handed barycentric frame: $\sref{B}$ for the chassis, $\sref{H}$ (and $\sref{K}$ in the second model) for the knuckles, and $\sref{W}$ for the wheels. Two additional frames, $\sref{G}$ (ground) and $\sref{S}$ (attached to the vehicle's centerline), are also defined. The frame $\sref{S}$ lies in the tangent plane, with its $x$-axis tangent to the centerline, pointing forward.
%
%Relative positions and orientations between frames are described by homogeneous transformation matrices \( g(d, \vPhi) \in SE(3) \), where \( d \in \mathbb{R}^3 \) is the Cartesian position and \( \vPhi \in \mathbb{R}^3 \) are the Euler angles parameterizing the rotation matrix \( R \in SO(3) \).
%
%The bodies are connected by joints, forming the kinematic tree shown in Fig.~\ref{fig:tree}. The first model is a special case of the second, where $\sref{H}$ and $\sref{K}$ are always coincident, so only the kinematic tree for the second model is presented to avoid redundancy.

%% END ORIGINAL

%The vehicle is modeled as an articulated rigid-body system. In particular, chassis, knuckles and wheels are considered as rigid bodies. The chassis comprises all the sprung masses, including the driver, which are rigidly attached to one another. The knuckles are assumed here as zero-mass bodies, and their function is to decouple the motion of the suspension due to steering and vertical travel to the wheel spindle. In the second model, the knuckles are split into two zero-mass bodies, introducing one additional degree of freedom (DoF) for each wheel that allows to simulate the active camber variation system, here named dynamic camber control (DCC). The wheels include all the rotating unsprung masses, while all non-rotating parts of the unsprung masses, such as the brake calipers are included in the wheel-tire mass.
%
%Each body has a right-handed barycentral reference frame attached. The frames are denoted as $\sref{B}$, $\sref{H}$ (and $\sref{K}$) and $\sref{W}$ for the chassis, knuckles and wheels, respectively, where $\sref{K}$ is present only in the second model. Two additional frames are introduced: $\sref{G}$, fixed to the ground, and $\sref{S}$, that follows the vehicle on the ground. More precisely, $\sref{S}$ is attached to the centerline curve with its $x$ and $y$ axes lying in the tangent plane, and the $x$ axis also tangent to the centerline curve pointing forward.
%
%The relative positions and orientations between reference frames are described by homogeneous transformation matrices $g(d,\vPhi)\in SE(3)$, where $d\in \bbR^3$ is the relative Cartesian coordinates vector and $\vPhi \in \bbR^3$ are the three Euler angles that parameterize the rotation matrix $R\in SO(3)$.
%
%The bodies are connected by joints and the kinematic tree of the second model is shown in Figure \ref{fig:tree}. It is worth noting that the first model can be seen as a particular case of the second, in which the reference frames $\sref{H}$ and $\sref{K}$ are constrained to be always overlapping. For this reason, and to avoid cluttering the document, has been chosen to show the kinematic tree only for the second model.

\begin{figure}[t]
	\centering
	\includegraphics[scale = 0.8]{Images/vehicle_frames}
	\caption{Reference frames: $\sref{G}$ is a fixed inertial reference frame with the $z$-axis antiparallel to the gravity vector; $\sref{S}$ is a reference frame following the vehicle on track; $\sref{B}$ and $\sref{K}$ are FLU (Forward Left Up) reference frames aligned with the principal axes of inertia of the chassis and wheel, respectively; $\sref{H}$ is an auxiliary reference frame necessary to add the camber DoF; and $\sref{W}$ rotates about the $y$-axis with respect to $\sref{K}$ by the wheel rotation angle.}
	\label{fig:frames}
\end{figure}

The joints modeled are: hub bearings, connecting the rims $\sref{W}$ to the knuckles $\sref{H}$ (or $\sref{K}$ for the second model), and the suspension linkages, connecting the knuckles $\sref{H}$ to the chassis $\sref{B}$. For the second model a 1 DoF revolute joint connecting the two parts of the knuckles is also present. The chassis is a floating-base and its position and orientation with respect to the ground are unconstrained. This situations is conveniently encoded by a virtual six-DoF joint connecting $\{B\}$ to $\{G\}$. The resulting systems have 4+4+6=14 and 4+4+4+6=18 degrees of freedom, respectively.
%After defining the system's topology, a set of suitable state variables needs to be defined to describe the configuration and motion of each body.

%To fully describe the position and orientation of the frame $\sref{B}$ w.r.t. $\sref{G}$ we define vector $q_{gb}=(d_{gb};\varPhi_{gb})$.
%Its first-order kinematics is encoded in twist $V_{gb}^b = (v_{gb};\omega_{gb})$. In general, a twist $V_{ij}^k$ represents the motion of the frame $\sref{J}$ w.r.t. the frame $\sref{I}$ in $\sref{K}$ components. Unless otherwise specified, we employ \emph{body} coordinates, therefore $k=j$; this choice aligns to ABA convention~\cite{Featherstone:book:2008}. For each of the 4 DoF's associated to the suspension motion, a variable $z$ (and its derivative $\dot{z}$) is defined representing spring's length variation with respect to the nominal value. With these definitions, along with variables $\delta$ and its derivatives, defined in Sec.\ref{sec:suspanal}, one can find the six-component vector $q_{bh}$ representing the pose of $\sref{H}$ w.r.t. $\sref{B}$ and the twist $V_{bh}$. Assuming the wheel is an axis-symmetrical rigid body, and given that the relative orientation of the frame $\sref{W}$ w.r.t. $\sref{H}$ (or $\sref{K}$) is not relevant for the scope, only the angular velocity $\omega$ is introduced. For the second model, it is also necessary to introduce the variable $q_{hk}$ to describe the relative position of the frame $\sref{K}$ w.r.t. $\sref{H}$.
\subsection{Suspension Analysis}
\label{sec:suspanal}
The suspensions connect the chassis to the unsprung masses. This connection cannot be described by an elementary joint, such as prismatic or revolute. Therefore, a set of six variables $q_{bh}$, that parameterize a homogeneous transformation matrix, is necessary. The transformation can be realized by composing three translations and three rotations (the ZXY Euler sequence was selected) using the global Product of Exponentials (PoE) formula~\cite{Murray:book:1994}, which reads: \begin{equation}\label{eqn:PoE}
	g_{bh}=e^{\hat{Y}_1q_{bh,1}}\cdots e^{\hat{Y}_6q_{bh,6}} g_{bh}(0),
\end{equation}
where the offset $g_{bh(0)}=I$ and $Y_i=e_i$, with $e_i\in\bbR^6$ are the six normalized screw vectors (canonical basis elements in $\bbR^6$) associated with the Cartesian coordinates and the Euler ZXY angles in $q_{bh}$.
Clearly, the six components of $q_{bh}$ are not independent; the modeled suspension is, in fact, a double wishbone that has: 2 DoFs for wheels that can steer, and 1 DoF otherwise.
In this case, both axles have wheels that can steer; hence, a pair of independent variables ($z;\delta $) is associated with each wheel.
A detailed analysis is omitted here for the sake of brevity. The interested reader is referred to~\cite{Domenighini:Designs:2023} for a similar treatment using the PoE formalism.

>>>>

The second approach plans over the full horizon while incorporating a feedback policy that mimics the driver's closed-loop behavior. This enables a more realistic propagation of the uncertainty encoded in the covariance matrix $\bP$, preventing its uncontrolled growth through the stabilizing effect of the driver's feedback.

The approach is based on the three following steps: i) A nominal time-optimal trajectory is planned: this will be referred to as nominal feed-forward optimal trajectory. This is the trajectory a driver should follow in the ideal situation of a world without uncertainty. Therefore, in this step the mean trajectory is the actual trajectory a deterministic vehicle dynamics would trace. ii) A closed-loop controller mimicking the driver's action is computed which stabilizes the nominal feed-forward optimal trajectory computed at step i). We assume a discrete time-varying Linear Quadratic Regulator (LQR) controller is a good approximation of how an expert driver can track a prescribed trajectory. iii) In the last step the stochastic framework is reintroduced and the time-optimal planning problem is robustified using the Lyapunov framework. The main idea is to \emph{re-plan} a mean trajectory of a stochastic system such that, a time-varying LQR controller that aims at stabilizing it, is able to properly tame the propagation of the uncertainty. This amounts to ensuring that the robustified path constraints are satisfied. More specifically, this implies two key aspects: first, that the covariance matrix is propagated accounting explicitly for the controlled system dynamics; and second, that the path constraints are robustly satisfied by incorporating the covariance estimation propagated through the closed-loop controlled system. Some aspects that may initially appear technical are in fact essential. In this third step, not only are states and controls re-planned, but also the time-varying LQR gains. In fact, they need to explicitly account for the changed requirements associated with stabilizing a trajectory different from the nominal one, which must now satisfy the robustified constraints.
\subsection{Suspension Analysis}
\label{sec:suspanal}
The suspensions connect the chassis to the unsprung masses. This connection cannot be described by an elementary joint, such as prismatic or revolute. Therefore, a set of six variables $q_{bh}$, that parameterize a homogeneous transformation matrix, is necessary. The transformation can be realized by composing three translations and three rotations (the ZXY Euler sequence was selected) using the global Product of Exponentials (PoE) formula~\cite{Murray:book:1994}, which reads: \begin{equation}\label{eqn:PoE}
	g_{bh}=e^{\hat{Y}_1q_{bh,1}}\cdots e^{\hat{Y}_6q_{bh,6}} g_{bh}(0),
\end{equation}
where the offset $g_{bh(0)}=I$ and $Y_i=e_i$, with $e_i\in\bbR^6$ are the six normalized screw vectors (canonical basis elements in $\bbR^6$) associated with the Cartesian coordinates and the Euler ZXY angles in $q_{bh}$.
Clearly, the six components of $q_{bh}$ are not independent; the modeled suspension is, in fact, a double wishbone that has: 2 DoFs for wheels that can steer, and 1 DoF otherwise.
In this case, both axles have wheels that can steer; hence, a pair of independent variables ($z;\delta $) is associated with each wheel.
A detailed analysis is omitted here for the sake of brevity. The interested reader is referred to~\cite{Domenighini:Designs:2023} for a similar treatment using the PoE formalism.
